\Title

### 2013-08-02 response_to_reviewers.txt

##############################################################################
# Reviewer comments:
##############################################################################
# Reviewer #1 (Comments for Author):
##############################################################################

>The text is clearly written and the written arguments are well spelled out.
>They observe that synonymous mutations in the C2-V5 region that disrupt
>modelled base pairs in SHAPE assays are more likely to be lost than other
>mutations, implying clear fitness costs to the virus, although they point out
>that those with major costs will never reach high frequencies. They argue that
>less deleterious mutations, but still with a fitness cost, can rise to high
>frequency by 'hitch hiking' with immune selected non-synonymous mutations, in
>other words simple linkage. The selection against synonymous slows down
>evolution of the virus so selective forces may have been underestimated in the
>past. These are important concepts and show that immune escape mutations
>should not be considered in isolation. They support the view that 'landscaping'
>approaches such as that described recently by Chakraborty in Immunity (Immunity
>38:606-617, 2013), need to be applied.

We thank the reviewer for this thoughtful assessment. We have included a
brief discussion of the landscape aspect of HIV fitness and now cite
Ferguson et al, Immunity, 2013.

##############################################################################
# Reviewer #2 (Comments for Author):
##############################################################################

>These are interesting results and the manuscript is clearly written. The second
>observation is not novel, but underappreciated, whereas this is the first
>concrete calculation of the average selective disadvantage of silent mutations
>in HIV taking draft into account. As such, it is necessary for this work to
>establish proper error estimates so that the community can make valid use of
>the results.  With this in mind, the manuscript can be strengthened in the
>following respects:

We thank the reviewer for the constructive comments, which have helped
us to improve the manuscript. We have resorted to a more sophisticated
sampling scheme for parameters in the computer simulations and now scan
a larger parameter space. All our conclusions remain valid, but are more
reliable now. We detail these improvements in our point to point
response below.

##############################################################################
>1) The modeling generation time of 1 day is odd.  It takes about a day for the
>virion to integrate and start transcription.  One usually takes 2 days as the
>mean generation time, but there is individual variation. See next point.

We agree that one generation in HIV takes closer to 2 days than 1
day. Most escape rates and selection coefficients, however, are more
readily measured in days and we have chosen to model HIV evolution with
generation of one day for simplicity. Factor of 2 changes in the
generation have next to now influence on the dynamics and only
renormalize genetic drift. Genetic drift is not a dominant force in HIV
evolution and the population size, the main determinant of drift, is not
known to this accuracy anyway. We have repeated a subset of the
simulations with a generation time of one day, finding now
difference. IS THIS SHOWN SOMEWHERE?

##############################################################################
>2) Related to the above: the simulation does not address how sensitive the
>results are to varying the parameters (in particular, the generation time, the
>outcrossing rate, the mutation rate, and the assigned deleterious fitness of
>the "non-synonymous" changes).  In the absence of described theoretical
>analysis shaping our expectations, I strongly feel that this sensitivity
>analysis needs to be part of the results, and folded into the error bars.

We have greatly extended the parameter space that we survey and now test
different values for the mutation rate, the recombination rate, the
generation time, the fitness effects, and the population size. To
accomodate this high dimensional data, we have changed figure 4. The
figure now shows the joint distribution of the deleterious effect and
the fraction of synonymous mutations that are deleterious that are
compatible with observations as a heat map. The marginal distribution of
the individual parameters consistent with the data are shown in the supplement.

The distribution of allowed deleterious is sharply peaked around
???. DISCUSS THE DIVERSITY ISSUES.

##############################################################################
>3) The reasonableness of the ranges for the various parameters should be argued
>with citation to existing literature.

We have added a discussion of the parameter ranges to the methods
section.

##############################################################################
>4) I am a bit surprised that the authors chose to do random sampling in their
>multidimensional parameter space.  One would have expected a Latin hypercube
>sampling to provide a better coverage.

Indeed. Our previous simulations where not limited by coverage of the
parameter space. Now that we cover a much larger (higher dimensional)
space, we did implement a Latin hypercube sampling scheme. We thank the
reviewer for this very useful suggestion. WE MIGHT WANT TO SHOW THE RANK
CORRELATION MATRIX BETWEEN THE DIFFERENT VARIABLES.

##############################################################################
>5) The actual dynamics of HIV in the body is presumably more complex: it is
>likely that the virions are produced in large bursts of virtually genetically
>identical virions most of which are then cleared by the body fast on the
>time-scale of HIV generations. There is no discussion whether this leads to
>increased stochasticity in the dynamics, nor whether this can influence the
>quantitative evaluations presented here. It is possible that the authors mean
>to absorb these into some kind of reduced effective population size, but a
>short discussion would be beneficial.

We agree with reviewer that HIV evolution is bursty, but we would like
to mention that actual number of cells that get infected every day is
around 10^7 and hence much larger than we are able to simulate. The
population size in our simulation can therefore be viewed as an
effective population size. We have investigated the sensitivity of our
results to changes in the population size and find that our conclusions
are insensitive to the population size as long as the product of
population size and mutation rate is not very much smaller than 1.

WE SHOULD ADD A DISCUSSION AND POINT OUT HERE.

##############################################################################
>6) In the discussions, the rise of synonymous mutations is attributed to draft,
>but the fall to selective effects. In general, some of the fall might also be
>due to overlapping selective sweeps: the manuscript does not discuss this
>possibility.

Yes, much of the stochastic component in the dynamics of the frequencies
of synonymous mutations will be due to draft. The directed component can
be attributed to selection. WE ADDED A DISCUSSION OF THIS ISSUE.

##############################################################################
>7) The observed neutral fixation pattern of the nonsynonymous mutations is
>mentioned, and two explanations have been proposed to duet of increased
>diversity over synonymous mutations with no increase of fixation over neutral
>expectations.  Neither of the explanations, however, are plausible (at least,
>without argument) absent experimental evidence not mentioned in the text. It
>would perhaps be better to either strengthen this discussion with citation to
>literature, or to weaken it by mentioning these as mere possibilities amongst
>others.

We agree with the reviewer that the both hypothesis we put forward to
explain the fixation pattern of non-synonymous mutations are somewhat
tentative. Qualitatively, however, the processes we discuss have been
observed and we do cite experimental evidence. NEED TO ELABORATE.

### 2013-08-02 synmut.tex
added a response to reviewers file, changed discussion of Landscape paper

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\graphicspath{{./figures/}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\comment}[1]{\textit{\textcolor{red}{#1}}}
\newcommand{\mut}{\mu}
\newcommand{\mfit}{\langle F\rangle}
\newcommand{\mexpfit}{\langle e^{F}\rangle}
\newcommand{\pfix}{P_{\mathrm{fix}}}
\newcommand{\ox}{r}
\newcommand{\co}{\rho}
\newcommand{\gt}{g}
\newcommand{\locus}{s}
\newcommand{\locuspm}{t}
\newcommand{\OO}{\mathcal{O}}
\newcommand{\FIG}[1]{Fig.~\ref{fig:#1}}
\newcommand{\FIGS}[2]{Figs.~\ref{fig:#1} and~\ref{fig:#2}}
\newcommand{\env}{\textit{env}}
\newcommand{\rev}{\textit{rev}}
\newcommand{\pol}{\textit{pol}}
\newcommand{\shankaregion}{C2-V5}
\newcommand{\PCApat}{1}
\newcommand{\syndiv}{2}
\newcommand{\timedependence}{3}
\newcommand{\withinepi}{4}
\newcommand{\gentimetwo}{5}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\renewcommand{\thesubfigure}{\Alph{subfigure}}
\newcommand{\Author}{Fabio~Zanini and Richard~A.~Neher}
\newcommand{\Title}{Quantifying selection against synonymous mutations in HIV-1 \env{} evolution}
\newcommand{\Affiliation}{Max Planck Institute for Developmental Biology, 72076 T\"ubingen, Germany}
\newcommand{\Keywords}{{HIV}, {synonymous}, {population genetics}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage{hyperref}
\hypersetup{colorlinks,linkcolor=red,citecolor=blue, pdfauthor={\Author}, pdftitle={\Title}, pdfkeywords={\Keywords}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\linespread{1.5}
\begin{document}
\title{\Title}
\author{\Author\\ \Affiliation}
%\affiliation{\Affiliation}
\date{\today}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\maketitle

\newpage
%\begin{abstract}
\section*{Abstract}
\noindent
Intrapatient HIV-1 evolution is driven by the adaptive immune system
resulting in rapid change of HIV-1 proteins. When cytotoxic CD8${}^+$ T-cells
or neutralizing antibodies target a new epitope, the virus often escapes via
nonsynonymous mutations that impair recognition. Synonymous mutations do not
affect this interplay and are often assumed to be neutral. We test this
assumption by tracking  synonymous mutations in 
longitudinal intrapatient data from the \shankaregion{} part of the \env{}
gene. We find that most synonymous variants are lost even though
they often reach high frequencies in the viral population, suggesting a
cost to the virus. Using published data
from SHAPE assays, we find that synonymous mutations that disrupt base pairs
in RNA stems flanking the variable loops of gp120 are more likely to be lost than other
synonymous changes: these RNA hairpins might be important for HIV-1.
Computational modeling indicates that (i) most synonymous mutations in this
genomic region have a selection coefficient of the order of $-0.004$, and
(ii) synonymous variants rise to high frequency by genetic hitchhiking on neighboring beneficial
variants. This weak selection against synonymous substitutions does not
result in a strong pattern of conservation in cross sectional data, but
slows down the rate of evolution considerably. Our findings are consistent with the
notion that large scale patterns of RNA structure are functionally
relevant, while the precise base pairing pattern is not.
%\end{abstract}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Introduction}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
HIV-1 evolves rapidly within a single host during the course of the infection.
This evolution is driven by strong selection imposed by the host immune system
via cytotoxic CD8${}^+$ T-cells (CTLs) and neutralizing antibodies
(nAbs)~\citep{rambaut_causes_2004} and facilitated by HIV-1's high mutation rate
~\citep{mansky_lower_1995,abram_nature_2010}. Escape mutations
in epitopes targeted by CTLs are typically observed during early infection and spread
rapidly through the population~\citep{mcmichael_immune_2009}. During chronic
infection, the most rapidly evolving parts of the HIV-1 genome are the variable
loops V1-V5 in the envelope protein gp120 (V loops), which change to avoid recognition by
nAbs. Escape mutations in \env{}, the gene encoding gp120, spread through the
viral population within a few months.
Consistent with this time scale, it is found that serum from a particular time
typically neutralizes autologous virus extracted more than 3-6 months earlier but not contemporary
virus \citep{richman_rapid_2003}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Computer simulations}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Computer simulations were performed using FFPopSim
\citep{zanini_ffpopsim:_2012}. Briefly, FFPopSim enables individual-based
simulations where each site in the genome is represented by one bit that can be
in one of two states. Outcrossing rates, crossover rates, mutations rates and
arbitrary fitness functions can be specified. We used an outcrossing rate of
$r=0.01$ per day \citep{batorsky_estimate_2011, neher_recombination_2010},
a mutation rate of $\mu=2 \times 10^{-5}$ \citep{mansky_lower_1995, abram_nature_2010}.
We performed the simulations with
generation times of both 1 and 2 days, because this parameter varies across patients
\citep{perelson_hiv-1_1996, perelson_decay_1997}; our conclusions are shown in
did not change (see \FIG{simheat} and \figurename~S\gentimetwo, respectively). We simulated intrapatient
evolution for 6000 days. For simplicity, third positions of every codon were
deemed synonymous and assigned either a selection coefficient $0$ with
probability $1-\alpha$ or a deleterious effect $s_d$ with probability $\alpha$.
Mutations at the first and second positions were assigned strongly deleterious 
fitness effects 0.02. At 
rate $k_A$, a random locus in the genome is designated an epitope that can
escape by one or several mutations with exponentially distributed escape rates
with mean $\epsilon$. Both full-length HIV-1 genomes and \env{}-only simulations
were performed and yielded comparable results.

The simulations were repeated 2400 times with random choices for the following
parameters: the fraction of deleterious sites $\alpha$ was sampled uniformly
between 0.75 and 1.0; the average deleterious effect $s_d$ was sampled such that
its logarithm was uniformly distributed  between $2 \cdot 10^{-4}$ and $2 \cdot 10^{-2}$; the
average escape rate $\epsilon$ of escape mutation was sampled such that its logarithm was
uniform between $2 \cdot 10^{-2.5}$ and $2 \cdot 10^{-1.5}$ and the rate $k_A$ of new antibody
challenges such that its logarithm was uniform between $10^{-3}$ and $10^{-2}$
per generation. Populations were initialized with a homogeneous founder
population and were kept at an average size of $N=10^4$ throughout the
simulation. After 30 generations of burn-in to create genetic diversity, new
epitopes were introduced at a constant rate $k_A$.

@@

The areas below or above the neutral fixation probability (diagonal line) were
estimated from the binned fixation probabilities using linear interpolation
between the bin centers. This measure is sufficiently precise for our purposes.
In 10 runs out of 2400, the highest frequency bin was empty so the fixation
probability could not be calculated; those runs were excluded from
\FIG{simsfig}.

@@

To map the parameter range of the model that is compatible with the data, we
repeatedly simulated the evolution with random choices for the
parameters, see \FIG{simsfig}. Among all simulations, we selected the ones
that show $A_\text{syn}$ and $P_\text{interm}$ as observed in the data, i.e., a
large depression in fixation probability of synonymous SNVs but, simultaneously,
a moderately high synonymous diversity. Specifically, \FIG{simsfig} shows
parameter combinations for which we found $A_\text{syn} < -0.15$ and $0.0025 <
P_\text{interm} < 0.010$. This subset of parameters indicates that a high fraction
($\gtrsim 0.8$) of sites has to be deleterious with effect size $|s_d| \sim
0.004$. The inverse of $|s_d|$ is of the same order of magnitude as the
fixation/extinction times estimated above (see \FIG{fixp1}).

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Discussion}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
By analyzing the fate of single nucleotide variants (SNVs) in
longitudinal data of HIV-1 \env{} evolution, we demonstrated selection
against synonymous substitutions in the relatively conserved regions
C2-C4. We suggest through computational modeling that these
SNVs have deleterious effects on the order of $0.004$ and that they are
brought to high frequency through linkage to adaptive mutations.
Comparison with biochemical data (SHAPE) of base pairing propensity in the RNA
genome of HIV-1 indicated that these mutations tend to disrupt RNA secondary structures\citep{watts_architecture_2009}. Computer models
of RNA folding predict stable hairpins in these regions that
have been suggested to be functional and termed ``insulating
stems'' \citep{watts_architecture_2009, sanjuan_interplay_2011}.
The weak selection against synonymous variants is compatible with the
negative results of \textit{In vitro} replication assays investigating
the fitness effects of small RNA hairpins in HIV-1
\citep{knoepfel_role_2013}: it would take hundreds of cell culture
passages to detect fitness effects of the order of one per mille. The longitudinal data,
however, spans many years, hence our analysis is able to quantify the
subtle fitness effect of RNA structure within single infections.

The fixation probabilities and the sojourn times of SNVs represent a rich and simple
summary statistics: they are useful to characterize longitudinal sequence data and
can be compared to models via computer simulations. These statistics are informative
even in absence of a neutral control and thus appropriate to analyze
properties of synonymous sites. 
We find selection against synonymous substitutions despite the
fact that the corresponding sites are not strongly conserved in
cross sectional data. This is consistent with a recent comparative
analysis of SIV and HIV-1 RNA secondary structure using SHAPE assays and
computational methods \citep{pollom_comparison_2013}. While large scale
patterns of RNA structures tend to agree in both viruses, the individual
base pairs forming the structures are almost always
discordant. Even though the molecular architecture of these structures
changes over time, selection seems to maintain them, which reduces the
fixation probability and hence the rate of evolution at synonymous
sites. As expected from this argument, the evolutionary rate
at synonymous sites varies greatly along the HIV-1 genome
\citep{mayrose_towards_2007} (see also \figurename~S\syndiv). This variation can confound estimates of
selection on proteins substantially \citep{ngandu_extensive_2008}. 
The dynamic nature of HIV secondary structure makes it hard to find the 
compensatory mutations that would restore base pairing in the
longitudinal data; in fact, the exact base pairing pattern is
most likely different than in the reference sequence.

@@

\begin{figure}
\begin{center}
\subfloat{\includegraphics[height=0.3\linewidth]{reactivities_histograms_syn.pdf}\label{fig:SHAPEA}}
\subfloat{\includegraphics[height=0.3\linewidth]{fixation_probabilities_VnonV.pdf}\label{fig:SHAPEB}}
\caption{Permissible synonymous mutations tend to be unpaired.
Panel A) shows the distribution of SHAPE reactivities among sites at which synonymous 
SNVs fixed (red), sites at which SNVs reached frequencies above 15\% but
were subsequently lost (blue), and sites at which no high-frequency SNVs were observed (green) 
(all categories are restricted to the regions V1-V5$\pm 100$bp).
Sites at which SNVs fixed tend to have higher SHAPE reactivities, corresponding to
less base pairing, than those at which SNVs are lost.
Sites at which no SNVs are observed show an intermediate distribution of SHAPE values.
Panel B) shows the fixation probability of synonymous SNVs in
\shankaregion{} separately for variable regions V3-V5 and the connecting conserved 
regions C2-C4 that harbor RNA stems. As expected, the fixation probability is lower
inside the conserved regions. Data from Refs.~\cite{shankarappa_consistent_1999,
bunnik_autologous_2008, liu_selection_2006}.}
\label{fig:SHAPE}
\end{center}
\end{figure}
\begin{figure}
\begin{center}
\subfloat{\includegraphics[width=0.6\linewidth]{simulations_graduallydel.pdf}
\label{fig:simfixpvar}}\\
\subfloat{\includegraphics[width=0.6\linewidth]{simulations_syn.pdf}
\label{fig:simsfig}}
\caption{Distribution of selection coefficients on synonymous sites. Panel A)
The depression in $\pfix$ depends on the deleterious effect size 
of synonymous SNVs. This parameter also reduces synonymous
diversity, measured by the probability of a SNV to be found at
intermediate frequencies $P_\text{interm}$ (first inset).
Panel B) To assess the parameter space that affects synonymous fixation and
diversity, we run 2400 simulations with random parameters for deleterious effect
size, fraction of deleterious synonymous sites, average escape rate $\epsilon$
(color, blue to red corresponds to $10^{-2.5}$ to $10^{-1.5}$ per day), and rate of
introduction of new epitopes (marker size, from $10^{-3}$ to $10^{-2}$ per
day). Only simulations that reproduce the synonymous diversity and fixation
patterns observed in data are shown. These simulations demonstrate that
deleterious effects are around $-0.004$ and a large fraction of the 
synonymous mutations needs to be deleterious. As expected, larger
$s_d$ require larger $\epsilon$. Parameters are chosen
from prior distributions uniform in logspace as indicated by the red rectangle
(see methods).}
\label{fig:simheat}
\end{center}
\end{figure}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\graphicspath{{./figures/}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\comment}[1]{\textit{\textcolor{red}{#1}}}
\newcommand{\mut}{\mu}
\newcommand{\mfit}{\langle F\rangle}
\newcommand{\mexpfit}{\langle e^{F}\rangle}
\newcommand{\pfix}{P_{\mathrm{fix}}}
\newcommand{\ox}{r}
\newcommand{\co}{\rho}
\newcommand{\gt}{g}
\newcommand{\locus}{s}
\newcommand{\locuspm}{t}
\newcommand{\OO}{\mathcal{O}}
\newcommand{\FIG}[1]{Fig.~\ref{fig:#1}}
\newcommand{\FIGS}[2]{Figs.~\ref{fig:#1} and~\ref{fig:#2}}
\newcommand{\env}{\textit{env}}
\newcommand{\rev}{\textit{rev}}
\newcommand{\pol}{\textit{pol}}
\newcommand{\shankaregion}{C2-V5}
\newcommand{\PCApat}{1}
\newcommand{\syndiv}{2}
\newcommand{\timedependence}{3}
\newcommand{\withinepi}{4}
\newcommand{\marginals}{5}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\renewcommand{\thesubfigure}{\Alph{subfigure}}
\newcommand{\Author}{Fabio~Zanini and Richard~A.~Neher}
\newcommand{\Title}{Quantifying selection against synonymous mutations in HIV-1 \env{} evolution}
\newcommand{\Affiliation}{Max Planck Institute for Developmental Biology, 72076 T\"ubingen, Germany}
\newcommand{\Keywords}{{HIV}, {synonymous}, {population genetics}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage{hyperref}
\hypersetup{colorlinks,linkcolor=red,citecolor=blue, pdfauthor={\Author}, pdftitle={\Title}, pdfkeywords={\Keywords}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\linespread{1.5}
\begin{document}
\title{\Title}
\author{\Author\\ \Affiliation}
\date{\today}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\maketitle

\newpage
\section*{Abstract}
\noindent
Intrapatient HIV-1 evolution is driven by the adaptive immune system
resulting in rapid change of HIV-1 proteins. When cytotoxic CD8${}^+$ T-cells
or neutralizing antibodies target a new epitope, the virus often escapes via
nonsynonymous mutations that impair recognition. Synonymous mutations do not
affect this interplay and are often assumed to be neutral. We test this
assumption by tracking  synonymous mutations in 
longitudinal intrapatient data from the \shankaregion{} part of the \env{}
gene. We find that most synonymous variants are lost even though
they often reach high frequencies in the viral population, suggesting a
cost to the virus. Using published data
from SHAPE assays, we find that synonymous mutations that disrupt base pairs
in RNA stems flanking the variable loops of gp120 are more likely to be lost than other
synonymous changes: these RNA hairpins might be important for HIV-1.
Computational modeling indicates that (i) most synonymous mutations in this
genomic region have a selection coefficient of the order of $-0.003$, and
(ii) synonymous variants rise to high frequency by genetic hitchhiking on neighboring beneficial
variants. This weak selection against synonymous substitutions does not
result in a strong pattern of conservation in cross sectional data, but
slows down the rate of evolution considerably. Our findings are consistent with the
notion that large scale patterns of RNA structure are functionally
relevant, while the precise base pairing pattern is not.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Introduction}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
HIV-1 evolves rapidly within a single host during the course of the infection.
This evolution is driven by strong selection imposed by the host immune system
via cytotoxic CD8${}^+$ T-cells (CTLs) and neutralizing antibodies
(nAbs)~\citep{rambaut_causes_2004} and facilitated by HIV-1's high mutation rate
~\citep{mansky_lower_1995,abram_nature_2010}. Escape mutations
in epitopes targeted by CTLs are typically observed during early infection and spread
rapidly through the population~\citep{mcmichael_immune_2009}. During chronic
infection, the most rapidly evolving parts of the HIV-1 genome are the variable
loops V1-V5 in the envelope protein gp120 (V loops), which change to avoid recognition by
nAbs. Escape mutations in \env{}, the gene encoding gp120, spread through the
viral population within a few months.
Consistent with this time scale, it is found that serum from a particular time
typically neutralizes autologous virus extracted more than 3-6 months earlier but not contemporary
virus \citep{richman_rapid_2003}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Computer simulations}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Computer simulations were performed using FFPopSim
\citep{zanini_ffpopsim:_2012}. Briefly, FFPopSim enables individual-based
simulations where each site in the genome is represented by one bit that can be
in one of two states. Outcrossing rates, crossover rates, mutations rates and
arbitrary fitness functions can be specified. We used an outcrossing rate of
$r=0.01$ per day \citep{batorsky_estimate_2011, neher_recombination_2010},
a mutation rate of $\mu=2 \times 10^{-5}$ \citep{mansky_lower_1995, abram_nature_2010}.
We performed the simulations with
generation times of both 1 and 2 days, because this parameter varies across patients
\citep{perelson_hiv-1_1996, perelson_decay_1997}; our conclusions are shown in
did not change. We simulated intrapatient
evolution for 6000 days. For simplicity, third positions of every codon were
deemed synonymous and assigned either a selection coefficient $0$ with
probability $1-\alpha$ or a deleterious effect $s_d$ with probability $\alpha$.
Mutations at the first and second positions were assigned strongly deleterious 
fitness effects 0.02. At 
rate $k_A$, a random locus in the genome is designated an epitope that can
escape by one or several mutations with exponentially distributed escape rates
with mean $\epsilon$. Both full-length HIV-1 genomes and \env{}-only simulations
were performed and yielded comparable results.

The simulations were repeated 1000 times with a four dimensional Latin hypercube sampling scheme.
For two parameters, ranges were chosen around estimates from the recent literature:
the average escape rate $\epsilon$ of escape mutation was sampled such that its logarithm was
uniform between $2 \cdot 10^{-2.5}$ and $2 \cdot 10^{-1.5}$
\citep{neher_recombination_2010} and the rate $k_A$ of new antibody
challenges such that its logarithm was uniform between $10^{-3}$ and $10^{-2}$
per generation \citep{richman_rapid_2003, moore_limited_2009}.
For the other two parameters, ranges were determined to include all simulations
that reproduced the experimental data: the fraction of deleterious sites $\alpha$ was sampled uniformly
between 0.75 and 1.0; the average deleterious effect $s_d$ was sampled such that
its logarithm was uniformly distributed  between $2 \cdot 10^{-4}$ and $2 \cdot
10^{-2}$. Populations were initialized with a homogeneous founder
population and were kept at an average size of $N=10^4$ throughout the
simulation. After 30 generations of burn-in to create genetic diversity, new
epitopes were introduced at a constant rate $k_A$.

@@

The areas below or above the neutral fixation probability (diagonal line) were
estimated from the binned fixation probabilities using linear interpolation
between the bin centers. This measure is sufficiently precise for our purposes.
In 4 runs out of 1000, the highest frequency bin was empty so the fixation
probability could not be calculated; those runs were excluded from
\FIG{simsfig}.

@@

To map the parameter range of the model that is compatible with the data, we
repeatedly simulated the evolution with random choices for the
parameters, see \FIG{simsfig}. Among all simulations, we selected the ones
that show $A_\text{syn}$ and $P_\text{interm}$ as observed in the data, i.e., a
large depression in fixation probability of synonymous SNVs but, simultaneously,
a moderately high synonymous diversity. Specifically, \FIG{simsfig} shows
parameter combinations for which we found $A_\text{syn} < -0.15$ and $0.0025 <
P_\text{interm} < 0.010$. This subset of parameters indicates that a high fraction
($\gtrsim 0.8$) of sites has to be deleterious with effect size $|s_d| \sim
0.003$. The inverse of $|s_d|$ is of the same order of magnitude as the
fixation/extinction times estimated above (see \FIG{fixp1}).

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Discussion}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
By analyzing the fate of single nucleotide variants (SNVs) in
longitudinal data of HIV-1 \env{} evolution, we demonstrated selection
against synonymous substitutions in the relatively conserved regions
C2-C4. We suggest through computational modeling that these
SNVs have deleterious effects on the order of $0.003$ and that they are
brought to high frequency through linkage to adaptive mutations.
Comparison with biochemical data (SHAPE) of base pairing propensity in the RNA
genome of HIV-1 indicated that these mutations tend to disrupt RNA secondary structures\citep{watts_architecture_2009}. Computer models
of RNA folding predict stable hairpins in these regions that
have been suggested to be functional and termed ``insulating
stems'' \citep{watts_architecture_2009, sanjuan_interplay_2011}.
The weak selection against synonymous variants is compatible with the
negative results of \textit{In vitro} replication assays investigating
the fitness effects of small RNA hairpins in HIV-1
\citep{knoepfel_role_2013}: it would take hundreds of cell culture
passages to detect fitness effects of the order of one per mille. The longitudinal data,
however, spans many years, hence our analysis is able to quantify the
subtle fitness effect of RNA structure within single infections.

The fixation probabilities and the sojourn times of SNVs represent a rich and simple
summary statistics: they are useful to characterize longitudinal sequence data and
can be compared to models via computer simulations. These statistics are informative
even in absence of a neutral control and thus appropriate to analyze
properties of synonymous sites. 
We find selection against synonymous substitutions despite the
fact that the corresponding sites are not strongly conserved in
cross sectional data. This is consistent with a recent comparative
analysis of SIV and HIV-1 RNA secondary structure using SHAPE assays and
computational methods \citep{pollom_comparison_2013}. While large scale
patterns of RNA structures tend to agree in both viruses, the individual
base pairs forming the structures are almost always
discordant. Even though the molecular architecture of these structures
changes over time, selection seems to maintain them, which reduces the
fixation probability and hence the rate of evolution at synonymous
sites. As expected from this argument, the evolutionary rate
at synonymous sites varies greatly along the HIV-1 genome
\citep{mayrose_towards_2007} (see also \figurename~S\syndiv). This variation can confound estimates of
selection on proteins substantially \citep{ngandu_extensive_2008}. 
The dynamic nature of HIV secondary structure makes it hard to find the 
compensatory mutations that would restore base pairing in the
longitudinal data; in fact, the exact base pairing pattern is
most likely different than in the reference sequence. Nevertheless, the
importance of secondary structure underscores the realization that viral
fitness is a complicated function (fitness landscape) in the
high-dimensional space of possible genotypes
\citep{ferguson_translating_2013}. We show, that this landscape is shaped not
only by interactions between amino-acids, but also by ubiquitous
interactions between nucleotides in RNA structures.

@@

\begin{figure}
\begin{center}
\subfloat{\includegraphics[height=0.3\linewidth]{reactivities_histograms_syn.pdf}\label{fig:SHAPEA}}
\subfloat{\includegraphics[height=0.3\linewidth]{fixation_probabilities_VnonV.pdf}\label{fig:SHAPEB}}
\caption{Permissible synonymous mutations tend to be unpaired.
Panel A) shows the distribution of SHAPE reactivities among sites at which synonymous 
SNVs fixed (red), sites at which SNVs reached frequencies above 15\% but
were subsequently lost (blue), and sites at which no high-frequency SNVs were observed (green) 
(all categories are restricted to the regions V1-V5$\pm 100$bp).
Sites at which SNVs fixed tend to have higher SHAPE reactivities, corresponding to
less base pairing, than those at which SNVs are lost.
Sites at which no SNVs are observed show an intermediate distribution of SHAPE values.
Panel B) shows the fixation probability of synonymous SNVs in
\shankaregion{} separately for variable regions V3-V5 and the connecting conserved 
regions C2-C4 that harbor RNA stems. As expected, the fixation probability is lower
inside the conserved regions. Data from Refs.~\cite{shankarappa_consistent_1999,
bunnik_autologous_2008, liu_selection_2006}.}
\label{fig:SHAPE}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
\subfloat{\includegraphics[width=0.6\linewidth]{simulations_graduallydel.pdf}
\label{fig:simfixpvar}}\\
\subfloat{\includegraphics[width=0.6\linewidth]{simulations_syn_latin_contour.pdf}
\label{fig:simsfig}}
\caption{Distribution of selection coefficients on synonymous sites. Panel A)
The depression in $\pfix$ depends on the deleterious effect size 
of synonymous SNVs. This parameter also reduces synonymous
diversity, measured by the probability of a SNV to be found at
intermediate frequencies $P_\text{interm}$ (first inset).
Panel B) To assess the parameter space that affects synonymous fixation and
diversity, we run 3000 simulations with random parameters for deleterious effect
size, fraction of deleterious synonymous sites, average escape rate $\epsilon$,
rate of introduction of new epitopes, population size, mutation rate, and
recombination rate (see methods). The density of simulations that reproduce
the synonymous diversity and fixation patterns observed in data are shown.
These simulations demonstrate that deleterious effects are around $-0.003$
and a large fraction of the synonymous mutations needs to be deleterious. This
picture is also confirmed by the analysis of the marginal distributions of the
single parameters (see \figurename~S\marginals).}
\label{fig:simheat}
\end{center}
\end{figure}
